Live Functional Programming with Typed Holes

### 2018-07-12 abstract.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

Live programming environments aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors. Consequently,    
live feedback presented to the programmer has gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
% In some cases, these gaps persist over a substantial length of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.

@@

This paper confronts this \emph{gap problem} from type-theoretic first principles by developing 
\emph{a dynamic semantics for incomplete functional programs}, 
starting from the static semantics for incomplete functional programs
developed by \citet{popl-paper}. 
%
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as membranes around static and dynamic type inconsistencies. 
%
Rather than aborting with an exception when evaluation encounters any of these holes as in
several existing systems, evaluation proceeds around the holes,
performing as much of the remaining computation as is possible and tracking the 
closure around each hole instance as it flows through the program. Editor services can report information from these hole closures 
to help the programmer decide how to fill the holes in the program. 
Hole closures also enable a \emph{fill-and-resume} feature that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics borrows machinery from both gradual type theory (to handle
type holes) and contextual modal type theory (which provides a
logical basis for tracking hole closures), combining these and developing additional machinery necessary 
to continue evaluation past the hole forms in a manner that admits important metatheoretic properties.

>>>>

Live programming environments aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors. Consequently,    
live feedback presented to the programmer has temporal or perceptive gaps, i.e. it flickers in and out or it goes stale whenever the program becomes incomplete. 
% In some cases, these gaps persist over a substantial length of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.

@@

This paper confronts this ``{gap problem}'' from type-theoretic first principles by developing 
\emph{a dynamic semantics for incomplete functional programs}, 
starting from the static semantics for incomplete functional programs
developed by \citet{popl-paper}. 
%
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as membranes around static and dynamic type inconsistencies. 
%
Rather than aborting with an exception when evaluation encounters any of these holes as in
several existing systems, evaluation proceeds around the holes,
performing as much of the remaining computation as is possible and tracking the 
closure around each hole instance as it flows through the program. Editor services can report information from these hole closures 
to help the programmer decide how to fill the holes in the program. 
Hole closures also enable a \emph{fill-and-resume} feature that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics borrows machinery from both gradual type theory (to handle
type holes) and contextual modal type theory (which provides a
logical basis for tracking hole closures), combining these and developing additional machinery necessary 
to continue evaluation past holes while maintaining important metatheoretic properties.

### 2018-07-12 examples-1.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer at all.
In response, the programmer might insert a temporary universal placeholder, e.g. by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design). This would cause typechecking to succeed. 
However, when the programming environment attempts to evaluate the program, 
evaluation would proceed only as far as the first \li{map} iteration, 
which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
% While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.

@@

% The result itself is not the only live feedback that \Hazel reports to the programmer.
% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
Going further, \Hazel helps programmers reason about the dynamic behavior of expressions bound to variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}. 
In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments. 
New to our approach are the values associated with bindings, which come from the environment associated with the selected hole instance in the result. 
We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. 
% An environment is a partial mapping of variables in scope to values. 
The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline), or cycle through all of the closures for the hole at the cursor by clicking the arrows, e.g. $\blacktriangleright$, or using hotkeys.

@@

Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}. So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid during class) has filled in the base case, and in the recursive case, has selected the head of the list as the pivot, partitioned the remainder of the list relative to the pivot and made the two recursive calls. A hole appears in the return position of the function body as the programmer contemplates how to combine the available variables to fill the hole with an appropriate list expression, as indicated by the type inspector in Fig.~\ref{fig:qsort-type-inspector}.

At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
the indeterminate result of this function application is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result. For example, in its initial state (Fig.~\ref{fig:qsort-sidebars}, left) it shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}. From this, the programmer can confirm (or the lecturer can visually point out) that 
the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming? math?}.

>>>>

Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback.
In response, the programmer might encode a hole by raising an exception. This would cause typechecking to succeed. 
However, 
evaluation would proceed only as far as the first \li{map} iteration, 
which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
% While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.

@@

% The result itself is not the only live feedback that \Hazel reports to the programmer.
% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
Going further, \Hazel helps programmers reason about the dynamic behavior of expressions bound to variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}. 
In all three states, the live context inspector displays the names and types of the variables that are in scope. 
New to our approach are the values associated with bindings, which come from the environment associated with the selected hole instance in the result. 
We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical connection detailed in Sec.~\ref{sec:calculus}). 
In this case, there are three instances of hole \li{1} in the result, numbered sequentially \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages} by \li{map}. 
% An environment is a partial mapping of variables in scope to values. 
The programmer can directly select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the purple outline in Fig.~\ref{fig:grades-cell-mockup}), or cycle through all of the closures for the hole at the cursor by clicking the arrows at the bottom of the context inspector, e.g. $\blacktriangleright$, or using hotkeys.

@@

Let us now consider a second more sophisticated example: an incomplete implementation of the recursive quicksort function, shown in Fig.~\ref{fig:qsort-example-code}. So far, the programmer (perhaps a student, or a lecturer using \Hazel as a presentational aid) has filled in the base case, and in the recursive case, partitioned the remainder of the list relative to the head, and made the two recursive calls. A hole appears in return position as the programmer contemplates how to fill the hole with an appropriate expression of list type, as indicated by the \emph{type inspector} in Fig.~\ref{fig:qsort-type-inspector}.

At the bottom of the cell in Fig.~\ref{fig:qsort-example-code}, the programmer has applied \li{qsort} to an example list. However, 
the indeterminate result of this function application is simply an instance of hole \li{1}, which serves only to confirm that evaluation went through the recursive case of \li{qsort}. 
More interesting is the live context inspector, shown in three states in Fig.~\ref{fig:qsort-sidebars}, which provides feedback about the values of the variables in scope at hole \li{1} from the the various instances of hole \li{1} that appear in the result, either immediately or within an outer closure. For example, in its initial state (Fig.~\ref{fig:qsort-sidebars}, left) it shows the closure at the instance of hole \li{1} that appears immediately in the result due to the outermost application of \li{qsort}. From this, the programmer can confirm (or the lecturer can visually point out) that 
the lists \li{smaller} and \li{bigger} computed by the call to \li{partition} are appropriately named, and observe that they are not yet themselves sorted\todo{are there studies to cite here about how concrete examples help people in programming? math?}.

### 2018-07-12 examples-2.tex
substantially expanded the related work
updates to sec 2
updated ill-typed quicksort example to use the same test as fig 2 and fixed some formatting errors
tweaks to the figures and text
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

\subsection{Example 3: Live Programming with Static Type Errors}

% \begin{subfigure}[t]{\textwidth}
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth,interpolate=false,valign=t]{images/qsort-type-error.png}
\caption{Example 3: Ill-Typed Quicksort}
\label{fig:qsort-type-error}
\vspace{-8px}
\end{figure}
% \end{subfigure}

The previous examples were incomplete 
because of \emph{missing} expressions.
%
Now, we discuss programs that are incomplete, 
and therefore conventionally meaningless, because of
\emph{type inconsistencies}. 
%
Let us return to the quicksort example just described, 
but assume that the programmer has filled in the previous hole
as shown in Fig.~\ref{fig:qsort-type-error}. (In Sec.~\ref{sec:resumption}, we discuss
how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.)

The programmer appears to be on the right track conceptually
in recognizing that the pivot needs to appear between the 
smaller and bigger elements. 
However, the types do not quite work out: the \li{@} operator here
performs list concatenation, but the pivot is an integer. 
Most compilers and editors will report a static error message, 
of varying utility, to the programmer in this case (and \Hazel 
follows suit, not shown). 
Our argument is that this should not cause all feedback about 
the dynamic behavior of the program to ``flicker out'' or ``go stale''. After all,
the error is localized and there is perfectly good code elsewhere 
in the program (if not nearby, then perhaps far away).

Our approach, following the prior work of \citet{popl-paper}, 
is to semantically internalize the ``red outline'' around
type inconsistencies as a \emph{non-empty hole}.
From the outside, a non-empty hole behaves statically much like an empty hole.
Our approach takes the same approach for the dynamic semantics,
i.e. evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
The semantics also associates an environment with each instance of a non-empty hole (not shown). 
The difference is that evaluation also proceeds inside the hole, so that 
feedback about an expression that might ``almost'' be correct is available. 
In this case, the result at the bottom of Fig.~\ref{fig:qsort-type-error}
reveals that the programmer is on the right track: the list elements 
(here, for a shorter example for the sake of exposition) are in the right order.
They simply have not been combined correctly.

>>>>

\subsection{Example 3: Live Programming with Static Type Errors}
\label{sec:static-errors}

% \begin{subfigure}[t]{\textwidth}
\begin{figure}
\centering
\includegraphics[width=\textwidth,interpolate=false,valign=t]{images/qsort-type-error-inset.png}
% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
\vspace{2px}
\caption{Example 3: Ill-Typed Quicksort}
\label{fig:qsort-type-error}
\vspace{-4px}
\end{figure}
% \end{subfigure}

The previous examples were incomplete 
because of \emph{missing} expressions.
%
Now, we discuss programs that are incomplete, 
and therefore conventionally meaningless, because of
\emph{type inconsistencies}. 
%
Let us return to the quicksort example just described, 
but assume that the programmer has filled in the previous hole
as shown in Fig.~\ref{fig:qsort-type-error}. In Sec.~\ref{sec:resumption}, we discuss
how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.

The programmer appears to be on the right track conceptually
in recognizing that the pivot needs to appear between the 
smaller and bigger elements. 
However, the types do not quite work out: the \li{@} operator here
performs list concatenation, but the pivot is an integer. 
Most compilers and editors will report a static error message
to the programmer in this case, and \Hazel 
follows suit in the type inspector (shown inset in Fig.~\ref{fig:qsort-type-error}). 
However, our argument is that the presence of a static type error should not cause all feedback about 
the dynamic behavior of the program to ``flicker out'' or ``go stale'' --
after all, there are perfectly meaningful parts of the program (both nearby
and far away from the error) 
whose dynamic behavior may be of interest, or even assist the programmer in fixing the type error.
% After all,
% the error is localized and there is perfectly good code elsewhere 
% in the program (if not nearby, then perhaps far away).

Our approach, following the prior work of \citet{popl-paper}, 
is to semantically internalize the ``red outline'' around
type inconsistencies, representing it as a \emph{non-empty hole}.
Evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
The semantics also associates an environment with each instance of a non-empty hole,
so we can use the live context inspector essentially as in Fig.~\ref{fig:qsort-sidebars} (not shown). 
The difference is that evaluation also proceeds inside the hole, so that 
feedback about the type-inconsistent expression, which might ``almost'' be correct, is available. 
In this case, the result at the bottom of Fig.~\ref{fig:qsort-type-error}
reveals that the programmer is on the right track: the list elements 
appear in the correct order.
They simply have not been combined correctly.

### 2018-07-12 examples-3.tex
substantially expanded the related work
tweaks to the figures and text
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

\subsection{Live Programming with Gradual Type Errors}

\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
programs that would otherwise be rejected by static type systems---either
because type inference cannot reconstruct a valid type assignment, or because
there may not be a unique valid type assignment at all.
%
In either case, gradual type systems allow types to contain \emph{unknown
types}, and partially unknown types can be used in contexts that require
different types, so long as they are \emph{consistent} (intuitively, equal
modulo the unknown parts).
%
Dynamic casts are then inserted to ensure that these remaining static
discrepancies are not violated at run-time.

>>>>

% !TEX root = hazelnut-dynamics.tex

\subsection{Example 4: Type Holes and Dynamic Type Errors}
\label{sec:dynamic-type-errors}

% \begin{subfigure}[t]{\textwidth}
\begin{figure}
\centering
\includegraphics[width=0.73\textwidth,interpolate=false,valign=t]{images/cast-errors.png}
% \includegraphics[width=0.28\textwidth,interpolate=false,valign=t]{images/type-inconsistency.png}
\caption{Example 4: Type Holes and Dynamic Type Errors}
\label{fig:cast-errors}
\vspace{-6px}
\end{figure}
% \end{subfigure}

% So far, we have only discussed incomplete programs where a hole appears within an expression.
In \Hazel, the program can also be incomplete because holes appear in types. 
\citet{popl-paper} confirmed that the literature on \emph{gradual type systems} \cite{Siek06a,DBLP:conf/snapl/SiekVCB15} is directly relevant to the problem of reasoning statically with type holes, by identifying the type hole with the unknown type. 
Unsurprisingly, then, it is also relevant to the problem of running 
programs with type holes. Indeed, that is the purpose of gradual typing: to be able to run programs that are not yet sufficiently annotated with types by inserting \emph{casts} only where necessary. As such, let us consider only a small synthetic example to demonstrate what is unique to our approach.

Fig.~\ref{fig:cast-errors} defines a simple function, \li{f}, of two arguments. 
The type annotation on the first line leaves the type of those arguments unknown. 
As such, the \Hazel type system, following the gradual typing approach,
allows the body of the function to use those two
arguments at any type at all (that is, the hole type is universally consistent). 
In this case, the first argument, \li{simple}, is used at one type, \li{bool}, 
and the second argument, \li{x}, is used at two different types in the two branches (perhaps because the programmer made a mistake), 
first as an \li{int} and second as a \li{string}  (we use \li{^} for string concatenation).
Although \Hazel supports only local type inference as of this writing, 
a system that uses ML-style type reconstruction to fill type holes statically, like GHC Haskell, would only be able to fill the first hole. Leaving the second hole unfilled is 
a parsimonious alternative to arbitrarily or heuristically\todo{cite error localization papers?}{} choosing one of the possibilities and marking the
other uses of \li{x} as ill-typed.

At the bottom of the cell in Fig.~\ref{fig:cast-errors}, we have three 
example applications of \li{f}. In this case, we simply tuple the results, but we could also
have made them into tests or put them into separate cells. All three are statically
well-typed, again because the hole type is universally consistent.
The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
of \li{f} is dynamically unproblematic. This might help the programmer confirm that at least
the first branch of the function body has been implemented as intended without the need to comprehensively
address all of the problems in the program. This flexibility is a common motivation for dynamic languages in programming folklore.

The second application of \li{f}, in contrast, causes a dynamic type error because the second argument, \li{2}, is an \li{int} but evaluation takes the branch where \li{x} is used as a \li{string}. 
Rather than aborting evaluation immediately when this occurs, as in existing gradual type systems, the problematic sub-term becomes a \emph{failed cast} term, shown shaded in red, which can be read ``\li{2} is an \li{int} that was used through a variable of hole type (\li{?}) as a \li{string}''. 
A failed cast behaves much like a non-empty hole in that it simply becomes an opaque value of the target type, \li{string}, that causes the surrounding concatenation operation to become indeterminate. 
Evaluation can continue on to the third application of \li{f}, which again is problematic, this time because the first argument is not a \li{bool} (perhaps because the programmer had an incorrect understanding of the argument order). 
Again, this causes a failed cast to appear, this time in guard position. Like a hole in guard position, evaluation cannot determine which branch to take so the whole guard expression becomes indeterminate. 
The pretty printer
hides the two branches behind ellipses.

In this synthetic example, it might have only been a small burden for the programmer to provide the intended types in the signature of \li{f}, but there are situations (e.g. during rapid prototyping or a live performance) where the programmer may consider the burden more substantial. This approach ensures that dynamic feedback does not have gaps even when there is a dynamic type error.

\subsection{Design Variations}

\begin{comment}
\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
programs that would otherwise be rejected by static type systems---either
because type inference cannot reconstruct a valid type assignment, or because
there may not be a unique valid type assignment at all.
%
In either case, gradual type systems allow types to contain \emph{unknown
types}, and partially unknown types can be used in contexts that require
different types, so long as they are \emph{consistent} (intuitively, equal
modulo the unknown parts).
%
Dynamic casts are then inserted to ensure that these remaining static
discrepancies are not violated at run-time.

### 2018-07-12 examples.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \cite{HazelnutSNAPL}. The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}), a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}, with which we assume basic  familiarity.
For the sake of 
exposition, we have ``post-processed'' the screenshots in this section to make use of a few language conveniences from \Elm~that were not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features demonstrated in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus in the supplement (see Sec.~\ref{sec:implementation}).

>>>>

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL} (initially for use by students and instructors in introductory functional programming courses). The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with code organized into cells and results appearing below each cell, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity.
For the sake of 
exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of some ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are orthogonal to the contributions of this paper. All of the live programming features demonstrated in the figures in this section have been implemented (see Sec.~\ref{sec:implementation}).

### 2018-07-12 fig-grades-example.tex
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

\begin{figure}[t]

\includegraphics[width=1.01\textwidth,interpolate=false]{images/grades-cell-mockup.png}

% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

@@

% \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}

\caption{Hazel mockups for Example 1a.}
\label{fig:grades-example}
\end{figure}

>>>>

\begin{figure}[t]
\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.8\textwidth,interpolate=false]{images/grades-cell-mockup.png}
\vspace{-3px}
\caption{Evaluating an incomplete functional program past the first hole}
\label{fig:grades-cell-mockup}
\end{subfigure}

\vspace{10px}

\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
~${}^\blacktriangleright$
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
~${}^\blacktriangleright$
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
\caption{The live context inspector communicates relevant static \emph{and} dynamic information about variables in scope.}
\label{fig:grades-sidebar}
\end{subfigure}
% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

@@

% \includegraphics[scale=0.20]{images/hazel-placeholder-1a.png}
\vspace{3px}

\caption{Example 1: Grades}
\label{fig:grades-example}
\vspace{-5px}
\end{figure}

### 2018-07-12 hazelnut-dynamics.tex
minor formatting for arxiv
fixed page numbers
arxiv version
fixed typos identified by OOPSLA reviews
tightened abstract
edits to sec 2 based on ravi's feedback, new example 4 with figure and a total rewrite of sec 2.4

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmsmall,review]{acmart}
\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}
\usepackage{xspace}
\usepackage{wrapfig}

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR}
}
\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

@@

\fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)

@@

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle
\thispagestyle{empty} % suppresses the footer

@@

\clearpage
\bibliography{references,all.short,hazel_NSF}

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
%\documentclass[acmsmall,review]{acmart}
%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
\documentclass[acmsmall,review,anonymous,nonacm]{acmart}
\settopmatter{printfolios=false,printccs=false,printacmref=false}
% \settopmatter{printccs=false,printacmref=false}

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}
\usepackage{xspace}
\usepackage{wrapfig}
% \usepackage[subtle]{savetrees}

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int, bool},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR}
}
\newcommand{\li}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}
\newcommand{\lismall}[1]{\lstinline[basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]{#1}}

@@

% \fancyfoot{} % suppresses the footer (also need \thispagestyle{empty} after \maketitle below)

@@

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle
% \thispagestyle{empty} % suppresses the footer

@@

%\clearpage
\bibliography{references,all.short,hazel_NSF}
